Composition-climate paper

### 2016-12-04 Composition_Ecology.Rmd

# Abstract
<!-- removed reference to 'higher Tmax' given results of sensitivity analyses.  Also took away the statement about 'reinforces shift to cooler conditions, again given ' 146 words after rewording-->
Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate relationships are poorly understood. We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land-use and climate change. Realized niches have shifted towards higher minimum temperatures and higher rainfall. A new attribution method implicates both historical climate change and land-use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces  shifts in species-climate relationships toward wetter distributions, or confounding, in which land use complicates shifts towards warmer distributions. Compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. These distributions of species in geographic and environmental space, combined with future climate scenarios, are used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventories offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
<!-- deleted Ordonez 2013 reference to shorten references -->
Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land-use, and other historical legacies [@svenning2015influence]. Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for slow ecological processes [@moorcroft2006close;@matthes2016benchmarking].

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions with a LOESS smooth superimposed for illustration. Modern climate (normals indicated by rectangles).  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. land-use data from the NLCD [@jin2013comprehensive].*")`
<!-- WICCI reference below:  Fix author in bibtex to WICCI, i.e. (WICCI, 2011) -->
Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land-use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land-use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.) [@barnes1979stem;@schlarbaum1998three;@castello1995pathogens].

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review these methods. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history]. At each survey point, surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid with an average of 61 survey points (122 trees) per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genus, and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in geographic or climate space of all cells with recorded presence for a specific taxon. <!-- inserted 'geographic or' because we often refer to range gains and losses in the paper -->

@@

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternatives for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data. We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplementary Information. There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatiotemporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].
<!-- Re-inserted paragraph below (I think it's important to explain why we picked the two modern normals that we did.  Note that there is a missing ref below, highlighted in bold.) -->
The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.  The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here. The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region (**REF**). We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.

@@

![](Final_Figures/figure_two.tiff)
<!-- Is Figure 2 for Larix or Quercus?  Legend here says Larix, later in text says Quercus -->
`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historical climate and vegetation change on observed shifts in tree-climate relationships. Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

@@

To assess whether the effects of historical land-use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different (Figure 2).   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If climate change is significant, but land-use is not significant, then the change is defined as confounding.  <!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed. -->

@@

Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C between the 1895-1924 and 1985-2014 normals (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C (see Supplementary Information for estimates based on the 1950 - 1979 normals). $P_{ann}$ increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.  Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al. 1985) and Fort Leavenworth, Kansas (Burnette et al. 2010). The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1). Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).

@@

Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all p < 0.0001).  The directionality of shifts is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ and $P_{ann}$ are higher (`r fig_nums("boxplot", display = "cite")`).  *Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while *Larix* has the smallest. Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm). These patterns are unchanged by a sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).

@@

`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs. changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use.  Results are plotted separately for each of the four climate variables considered here. Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig. 2).  For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`
<!-- Does Fig. 2 show Larix or Quercus?  This paragraph states Quercus; legend states Larix -->
Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).  The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (`r fig_nums("fourPanel", display = "cite")`).  For *Quercus* we see changes due to both climate (`r fig_nums("fourPanel", display = "cite")`b to d) and due to vegetation change (`r fig_nums("fourPanel", display = "cite")`b to c), but a stronger apparent shift for historical climate change, reflected in the attribution index $\Delta_{d} (`r fig_nums("hellPlot", display = "cite")`a).

<!-- Note that figure numbering is still off - Fig 6 should be Fig 5.  Needs fixing prior to submission-->
Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on $\Delta_{d} (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa. The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen precipitation niche shifts that are slightly more attributable to land-use than climate shifts (`r fig_nums("hellPlot", display = "cite")`a). The greatest land-use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$). The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land-use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land-use. The $T_{min}$ shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.

@@

```
<!-- Given new Table 1, have rewritten to state that all climate-veg interaction are compounding or confounding -->
<!-- Last sentence in para may need tweaking, if Supp Table 4 is changed - see comments there. -->
All climate-vegetation interactions are compounding ($\\Box$; n = 30) or confounding ($-$; n = 30), with no counteracting effects found (`r tab_nums("compTable", display = "cite")`). Compounding interactions, where both climate and land-use shifts push species distributions in the same direction along the climate gradient, are strongest for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space ($-_c$). Compounding interactions are also strong for $P_{ann}$ and are caused by both land-use and climate change shifting species to a wetter portion of climate space ($+_c$) than they occupied historically.  The patterns of confounding and compounding interactions are insensitive to alternative estimates of 19^th^ century temperature changes (Supplementary Information Table 3).  However, the apparent $T_{max}$ cooling and number of compounding interactions are reduced when 1950 - 1980 climate normals are used instead of 1985-2014 normals (Supplementary Table 4).

@@

<!-- Fix the WICCI ref, as above. -->
Compounding and confounding interactions between historical climate change and land-use have important implications for the management and genetics of forest communities. Some tree genera may have more resilience to climate change than expected based on contemporary observations alone. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 with $P_{ann}$ projections ranging from -17 to 110mm from a 1980s baseline [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates (primarily attributable to land use change, Figure 5), and $P_{ann}$ from 42 to 86mm lower than modern estimates (primarily attributable to climate change, Figure 5). The higher temperatures and lower precipitation associated with pre-settlement tree distributions may suggest greater resilience to climate change than might be assessed using modern data only.  This evidence for historical resilience to warmer and dryer climates is particularly critical for species that reach their range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).

@@

Acknowledgements
---------------------
The authors would like to thank the large number of individuals who have worked to undertake the Public Lands System Survey, to compile the original surveyor notes, to digitize and standardize the survey results, and finally, to assist in interpreting and compiling the data in its present form. In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group. The authors thank Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript. This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.  Maps were made using data from Natural Earth (http://naturalearth.com).

References
---------------

>>>>

# Abstract

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate relationships are poorly understood. We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land-use and climate change. Realized niches have shifted towards higher minimum temperatures and higher rainfall. A new attribution method implicates both historical climate change and land-use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces  shifts in species-climate relationships toward wetter distributions, or confounding, in which land use complicates shifts towards warmer distributions. Compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. These distributions of species in geographic and environmental space, combined with future climate scenarios, are used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventories offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land-use, and other historical legacies [@svenning2015influence]. Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for slow ecological processes [@moorcroft2006close;@matthes2016benchmarking].

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions with a LOESS smooth superimposed for illustration. Modern climate (normals indicated by rectangles).  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. land-use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land-use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land-use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.) [@barnes1979stem;@schlarbaum1998three;@castello1995pathogens].

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review these methods. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history]. At each survey point, surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid with an average of 61 survey points (122 trees) per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genus, and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in geographic or climate space of all cells with recorded presence for a specific taxon.

@@

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternatives for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data. We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplementary Information. There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatiotemporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].

The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.  The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here. The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region (FIA DataMart: http://apps.fs.fed.us/fiadb-downloads/datamart.html, accessed 1/12/2016). We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.

@@

![](Final_Figures/figure_two.tiff)

`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historical climate and vegetation change on observed shifts in tree-climate relationships. Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Quercus* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

@@

To assess whether the effects of historical land-use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different (Figure 2).   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If climate change is significant, but land-use is not significant, then the change is defined as confounding, as is the converse, since in both cases, the climate signal is weakened because of vegetation change.

@@

Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C between the 1895-1924 and 1985-2014 normals (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C (see Supplementary Information for estimates based on the 1950 - 1979 normals). $P_{ann}$ increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward ($r$ = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.  Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al. 1985) and Fort Leavenworth, Kansas (Burnette et al. 2010). The latter shows faster increases during winter (0.10^o^C decade^-1^) than summer (0.07^o^C decade^-1^). Hence, our estimates of historic climate change may be too conservative for $T_{min}$ and $T_{diff}$ and too large for $T_{max}$ (Supplementary Information).

@@

Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all $p$ < 0.0001).  The directionality of shifts is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ and $P_{ann}$ are higher (`r fig_nums("boxplot", display = "cite")`).  *Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{T_{min}}$ = 4.2 ^o^C), while *Larix* has the smallest. Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{T_{max}}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{T_{max}}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{P_{ann}}$ = 86mm), and *Larix* the smallest ($\delta_{P_{ann}}$ = 42mm). These patterns are unchanged by a sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).

@@

`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs. changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use.  Results are plotted separately for each of the four climate variables considered here. Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig. 2).  For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`

Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).  The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (`r fig_nums("fourPanel", display = "cite")`).  For *Quercus* we see changes due to both climate (`r fig_nums("fourPanel", display = "cite")`b to d) and due to vegetation change (`r fig_nums("fourPanel", display = "cite")`b to c), but a stronger apparent shift for historical climate change, reflected in the attribution index $\Delta_{d} (`r fig_nums("hellPlot", display = "cite")`a).

Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on $\Delta_{d} (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa. The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen precipitation niche shifts that are slightly more attributable to land-use than climate shifts (`r fig_nums("hellPlot", display = "cite")`a). The greatest land-use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$). The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land-use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land-use. The $T_{min}$ shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.

@@

All climate-vegetation interactions are compounding ($\\Box$; n = 30) or confounding ($-$; n = 30), with no counteracting effects found (`r tab_nums("compTable", display = "cite")`). Compounding interactions, where both climate and land-use shifts push species distributions in the same direction along the climate gradient, are strongest for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space ($-_c$). Compounding interactions are also strong for $P_{ann}$ and are caused by both land-use and climate change shifting species to a wetter portion of climate space ($+_c$) than they occupied historically.  The patterns of confounding and compounding interactions are insensitive to alternative estimates of 19^th^ century temperature changes (Supplementary Information Table 3).  However, the 1950 - 1980 climate normal results in non-significant $T_{max}$ changes, resulting in a loss of all compounding interactions for $T_{max}$ (Supplementary Table 4).  All $T_{max}$ interactions are then confounding, with significant shifts for vegetation changes (all positive), except for *Pinus* and *Thuja/Juniperus*, where changes of vegetation in climate space are non-significant.

@@

Compounding and confounding interactions between historical climate change and land-use have important implications for the management and genetics of forest communities. Some tree genera may have more resilience to climate change than expected based on contemporary observations alone. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 with $P_{ann}$ projections ranging from -17 to 110mm from a 1980s baseline [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates (primarily attributable to land use change, Figure 5), and $P_{ann}$ from 42 to 86mm lower than modern estimates (primarily attributable to climate change, Figure 5). The higher temperatures and lower precipitation associated with pre-settlement tree distributions may suggest greater resilience to climate change than might be assessed using modern data only.  This evidence for historical resilience to warmer and dryer climates is particularly critical for species that reach their range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).

@@

# Acknowledgements

The authors would like to thank the large number of individuals who have worked to undertake the Public Lands System Survey, to compile the original surveyor notes, to digitize and standardize the survey results, and finally, to assist in interpreting and compiling the data in its present form. In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group. The authors thank Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript. This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.  Maps were made using data from Natural Earth (http://naturalearth.com).

# References
